\documentclass[manuscript,nonacm]{acmart}
\usepackage[switch]{lineno}
%% acmart already provides amsmath, amssymb, amsthm, booktabs, xcolor, hyperref, graphicx.
\usepackage{mathtools}
\usepackage{stmaryrd}      % \llbracket, \rrbracket
\usepackage{mathpartir}    % \inferrule for SOS rules
\usepackage{listings}
\usepackage{tikz}
\usepackage[switch]{lineno}
\usetikzlibrary{fit,positioning,arrows.meta}
%% acmart predefines 'definition'; it does not define 'remark', so we add it.
\theoremstyle{acmdefinition}
\newtheorem{remark}{Remark}

%% ---- notation ----
\newcommand{\Loc}{\mathit{Loc}}
\newcommand{\Var}{\mathit{Var}}
\newcommand{\Val}{\mathit{Val}}
\newcommand{\TrId}{\mathit{TxId}}   % transaction identifiers
\newcommand{\Tid}{\mathit{Tid}}
\newcommand{\ThId}{\mathit{ThId}}   % thread identifiers (the scheduled thread i)
\newcommand{\TxId}{\mathit{TxId}}   % transaction identifiers (the live transaction t)
\newcommand{\Cmd}{\mathit{Cmd}}
\newcommand{\Expr}{\mathit{Expr}}
\newcommand{\store}{\sigma}
\renewcommand{\log}{\lambda}
\newcommand{\reduces}{\longrightarrow}
\newcommand{\tred}{\mathrel{\longrightarrow_{t}}}
\newcommand{\unit}{()}
\newcommand{\bbegin}{\textsf{begin}}
\newcommand{\eend}{\textsf{end}}
\newcommand{\retry}{\textsf{retry}}
\newcommand{\fork}{\textsf{fork}}
\newcommand{\getc}[1]{\textsf{get}\;(#1)}
\newcommand{\setc}[2]{\textsf{set}\,(#1,#2)}
\newcommand{\nin}{\bot}                 % "no transaction" marker
\newcommand{\subst}[3]{#1[#2\mapsto #3]}
\newcommand{\rulename}[1]{\textsc{\small #1}}

%% ---- evaluation-context notation ----

\newcommand{\ectx}{\mathsf{ECtx}}
\newcommand{\ehole}[1]{\ectx[[\, #1 \,]]}
%% runtime get/set actions: they record the value read/written and the
%% transaction token t, written get(l,v)@t and set(l,v)@t.
\newcommand{\getev}[3]{\textsf{get}(#1,#2)\,@\,#3}
\newcommand{\setev}[3]{\textsf{set}(#1,#2)\,@\,#3}
\newcommand{\elided}{\mathrm{actions}}

%% ---- abbreviations used in the worked-examples section ----
\newcommand{\st}[1]{[\,l\,{\mapsto}\,#1\,]}          % store over the single location l
\newcommand{\emap}[1]{\{#1\}}                          % thread map E
\newcommand{\adm}{\xrightarrow{\;\mathrm{adm}\;}}     % administrative reduction step
\newcommand{\ladm}[1][]{\ell_{\mathrm{adm}#1}}       % administrative-step label l_adm (opt. index, e.g. \ladm[,1])
\newcommand{\Abody}{\textsf{let}\;x\,{=}\,\getc{l}\;\textsf{in}\;\setc{l}{x\,{+}\,1}}
\newcommand{\Bbody}{\textsf{let}\;y\,{=}\,\getc{l}\;\textsf{in}\;
   (\textsf{if}\;y\,\texttt{==}\,0\;\textsf{then}\;(\setc{l}{99}\,;\,\retry)\;
   \textsf{else}\;\setc{l}{y\,{+}\,1})}

%% ---- declarative-semantics notation (Cerone et al. style) ----
\newcommand{\Op}{\mathit{Op}}
\newcommand{\EventId}{\mathit{EventId}}
\newcommand{\HEvent}{\mathit{HEvent}}
\newcommand{\WEvent}{\mathit{WEvent}}
\newcommand{\REvent}{\mathit{REvent}}
\newcommand{\rdev}[2]{\mathsf{get}(#1,#2)}   % our "read" is a get(x,n)
\newcommand{\wrev}[2]{\mathsf{set}(#1,#2)}   % our "write" is a set(x,n)
\newcommand{\retryev}{\rho}                 % retry event marker
\newcommand{\po}{\mathit{po}}
\newcommand{\maxpo}{\mathrm{max}_{\po}}
\newcommand{\minpo}{\mathrm{min}_{\po}}
\newcommand{\att}{\mathit{att}}
\newcommand{\anyop}[2]{\mathord{\_}(#1,#2)} % get(x,n) or set(x,n)
\newcommand{\pob}{\xrightarrow{\po}}        % program-order edge
\newcommand{\fst}{\mathrm{fst}}             % first projection of a history pair
\newcommand{\snd}{\mathrm{snd}}             % second projection of a history pair
\newcommand{\vs}[1]{\textcolor{blue}{\textbf{[VS:} #1\textbf{]}}}
%\renewcommand{\vs}[1]{}

%% ---- listings style for the Reagent language ----
\definecolor{kw}{rgb}{0.10,0.10,0.55}
\definecolor{cm}{rgb}{0.30,0.45,0.30}
\lstdefinelanguage{reagent}{
  keywords={let,in,get,set,begin,end,retry,if,then,else},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[n]{(*}{*)},
  morestring=[b]",
}
\lstset{
  language=reagent,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{kw}\bfseries,
  commentstyle=\color{cm}\itshape,
  showstringspaces=false,
  columns=fullflexible,
  frame=single,
  framesep=4pt,
  xleftmargin=4pt,
}
\newcommand{\code}[1]{\lstinline[basicstyle=\ttfamily]{#1}}

\title{The Reagent Language}

\author{Durwasa Chakraborty}
\email{durwasa.chakraborty@gmail.com}
\affiliation{%
  \institution{TODO: institution}
  \country{India}
}

\keywords{concurrency, transactional memory, operational semantics, internal consistency}
\begin{document}
\linenumbers
\begin{abstract}
TODO
\end{abstract}

\maketitle

\section{Abstract syntax}

\[
\begin{array}{rcll}
n     & \in & \mathbb{N}        & \text{natural-number literals}\\
l     & \in & \Loc              & \text{shared mutable locations}\\
x     & \in & \Var              & \text{immutable, let-bound variables}\\[4pt]
v     & ::= & n \mid \unit \mid \textsf{true} \mid \textsf{false} & \text{values (naturals, unit, booleans)}\\[4pt]
c     & ::= & \getc{l}          & \text{read a shared location}\\
      & \mid & \setc{l}{e}      & \text{write a shared location}\\
      & \mid & \bbegin          & \text{open a transaction}\\
      & \mid & \eend            & \text{commit the current transaction}\\
      & \mid & \retry           & \text{abort and restart the transaction}\\[4pt]
e     & ::= & v \mid x          & \text{values, variables}\\
      & \mid & e_1 + e_2        & \text{addition (the whiteboard's }\oplus\text{)}\\
      & \mid & \textsf{let}\;x = e_1\;\textsf{in}\;e_2 & \text{binding}\\
      & \mid & c                & \text{a command (and not a control-flow)}\\
      & \mid & \textsf{if}\;e\;\textsf{then}\;e\;\textsf{else}\;e & \text{conditional}\\
\end{array}
\]



\subsection{Semantics :: of the Reagent Language}
\label{subsec:semantics}

Each configuration $C$ in $\Phi$ represents the global state of the program at a particular point during execution. Thread identifiers form a fixed finite set $\ThId \subseteq \mathbb{N}$, and transaction identifiers a set $\TrId = \{t_1, t_2, \dots\}$. Formally, a configuration is a six-tuple $C = \langle\sigma,\lambda,\theta,E,\hat{\sigma},\hat{E}\rangle$, where:

\begin{itemize}

\item \(\sigma : \Loc \rightarrow \mathbb{N}\) maps each shared memory location to its current value.

\item \(\lambda \in \ThId \cup \{\bot\}\) records the \emph{thread} that currently holds the lock. $\lambda = i$ means thread $i$ owns the live transaction and is the only thread permitted to take a transactional step; $\lambda = \bot$ means no thread is active.

\item \(\theta \in \TrId \cup \{\bot\}\) is the identifier of the live transaction, minted \emph{fresh} by each $\bbegin$; $\theta = \bot$ means no transaction is active. It supplies the tag on the emitted actions --- when $\theta = t$ the labels read $\getev{l}{v}{t}$, $\eend\,@\,t$, and so on.

\item $E : \ThId \rightarrow \Expr$ maps each thread to the expression it is currently evaluating.

\item $\hat{\sigma}: \Loc \rightarrow \mathbb{N}$ is the snapshot of the store taken when the live transaction began, used to roll the store back on restart.

\item $\hat{E} \in \Expr$ is the expression from which execution resumes when a retry is encountered.

\end{itemize}

The snapshot $\hat{\sigma}$ and saved body $\hat{E}$ matter only inside a transaction, but we do not tag them with ``$\lambda \neq \bot$'' each time: an initial configuration already gives them values (Section~\ref{sec:init-final}) --- a run starts with $\lambda = \theta = \bot$ (no thread, no transaction) and $\hat{\sigma} = \sigma_0$, $\hat{E} = ()$ --- and every rule keeps them defined thereafter.

In every rule the \emph{active} thread is $i$ (the one taking the step) with
$E(i)=\ehole{r}$, where $r$ is the unique redex its evaluation context $\ectx$
locates (the subterm currently able to step). There is no scheduler: the
nondeterminism is inherent in $\rightarrow$, and any thread with an enabled redex
may be the one that steps. Every transactional rule
($\textsc{Get},\textsc{Set},\textsc{End},\textsc{Retry}$) carries the premises
$\lambda = i$ and $\theta = t$: \emph{only the thread that owns the live transaction
may step inside it}. Since $i \in \ThId$ and $t \in \TrId$, these premises imply
$\lambda \neq \bot$ and $\theta \neq \bot$ --- a transaction is live --- so the
snapshot $\hat{\sigma}$ and saved body $\hat{E}$ are always defined when such a rule
fires. The conclusion rewrites that redex in place and \emph{frames} the rest of the
table, $E' = E[i \mapsto \ehole{\cdot}]$ with $E'(j) = E(j)$ for every $j \neq i$;
the label above the arrow is tagged with the live transaction id $t$ (the value of
$\theta$).

\begin{figure}[ht]
\small
\begin{mathpar}
\inferrule*[left=\textsc{Get}]{
  E(i)=\ehole{\getc{l}} \\ \lambda = i \\ \theta = t \\ \sigma(l)=v
}{
  \langle\sigma,\lambda,\theta,E,\hat{\sigma},\hat{E}\rangle
  \xrightarrow{\getev{l}{v}{t}}
  \langle\sigma,\lambda,\theta,E[i\mapsto\ehole{v}],\hat{\sigma},\hat{E}\rangle
}
\and
\inferrule*[left=\textsc{Set}]{
  E(i)=\ehole{\setc{l}{v}} \\ \lambda = i \\ \theta = t
}{
  \langle\sigma,\lambda,\theta,E,\hat{\sigma},\hat{E}\rangle
  \xrightarrow{\setev{l}{v}{t}}
  \langle\sigma[l\mapsto v],\lambda,\theta,E[i\mapsto\ehole{\unit}],\hat{\sigma},\hat{E}\rangle
}
\and
\inferrule*[left=\textsc{Begin}]{
  E(i)=\ehole{\bbegin\,;\,e}
  \\
  \lambda=\bot
  \\
  t_{fresh}
}{
  \langle\sigma,\bot,\bot,E,\hat{\sigma},\hat{E}\rangle
  \xrightarrow{\bbegin\,@\,t_{fresh}}
  \langle
    \sigma,
    i,
    t_{fresh},
    E[i\mapsto\ehole{\unit}],
    \sigma,
    e
  \rangle
}
\and
\inferrule*[left=\textsc{End}]{
  E(i)=\ehole{\eend} \\ \lambda = i \\ \theta = t
}{
  \langle\sigma,\lambda,\theta,E,\hat{\sigma},\hat{E}\rangle
  \xrightarrow{\eend\,@\,t}
  \langle\sigma,\bot,\bot,E[i\mapsto\ehole{\unit}],\hat{\sigma},\hat{E}\rangle
}
\and
\inferrule*[left=\textsc{Retry}]{
  E(i)=\ehole{\retry} \\ \lambda = i \\ \theta = t
}{
  \langle\sigma,\lambda,\theta,E,\hat{\sigma},\hat{E}\rangle
  \xrightarrow{\retry\,@\,t}
  \langle
    \hat{\sigma},
    \bot,
    \bot,
    E[i\mapsto\ehole{\bbegin\,;\,\hat{E}}],
    \hat{\sigma},
    \hat{E}
  \rangle
}
\end{mathpar}
\caption{Semantics of Reagent commands. Each rule fires on the redex of the
active thread $i$; the transactional rules require $\lambda = i$ (only the
owner thread may step), and the emitted label is tagged with the live transaction
id $t$ (bound by $\theta = t$; $\textsc{Begin}$ mints it fresh). Unlisted six-tuple
fields are unchanged.}
\label{fig:semantics}
\end{figure}
Figure~\ref{fig:semantics} gives a formal description of the transition
rules. $\textrm{\sc{Get}}$ reads the value stored at a shared location
$l$. If the active thread $i$ owns the live transaction ($\lambda = i$) and $\sigma(l)=v$, then $\getc{l}$ returns the value $v$; the emitted action $\getev{l}{v}{t}$ (where $\theta = t$) records both the value $v$ read and the id $t$ of the live transaction. The store and transaction state remain unchanged.

$\textrm{\sc{Set}}$ updates a shared location. When the owner thread $i$ (with $\lambda = i$) executes
$\setc{l}{v}$, the value stored at location $l$ is updated to $v$ and the
command returns the unit value $\unit$; the emitted action $\setev{l}{v}{t}$ records the written value $v$ and the transaction id $t$.

$\textrm{\sc{Begin}}$ starts a transaction. No transaction may currently be alive ($\lambda = \bot$); transactions are therefore mutually exclusive, with at most one live transaction across all threads at any time. $\textrm{\sc{Begin}}$ mints a \emph{fresh} transaction id $t$, stores it in $\theta$ as the new live id, records the executing thread as the owner by setting $\lambda := i$, and saves the current store and continuation in $\hat{\sigma}$ and $\hat{E}$ respectively.

$\textsc{End}$ commits the current transaction, resetting both the owner $\lambda$ and the live-transaction register $\theta$ to $\bot$. The store is unchanged.


$\textsc{Retry}$ is retrying with \emph{lock-releasing}. Instead of keeping the thread inside the transaction and spinning, it (i) rolls the store back to the snapshot $\hat{\sigma}$, so the abandoned attempt has no effect; (ii) sets $\lambda=\bot$ and clears the live transaction id, \emph{releasing the single critical section} so that another thread may now $\bbegin$ and run; and (iii) re-arms $E(i)$'s expression as $\bbegin\,;\,\hat{E}$, so that when $i$ next steps it must re-acquire the lock through $\textsc{Begin}$ before re-running the saved body.

Because every transactional rule requires $\lambda = i$, a thread can only read, write, commit, or retry inside a transaction it opened itself. A $\getc{l}$, $\setc{l}{e}$, $\eend$, or $\retry$ reached with $\lambda \neq i$ --- in particular one with no matching enclosing $\bbegin$ --- has no rule to fire, so the run gets stuck rather than taking an ill-formed step. Well-formedness is thus enforced by the operational semantics itself, not assumed (Section~\ref{sec:examples}).

\subsection{Evaluation contexts}

An evaluation context $\ectx$ is an expression containing a single
hole $[\cdot]$ that marks exactly where the next reduction step occurs. We give
the evaluation context its own metavariable $\ectx$ precisely to keep it
separate from $E$, which represents a mapping from thread identifiers
($\mathit{Tid}$) to expressions ($\mathit{Expr}$); the two are distinct objects
and should not be conflated.
\[
\ectx \;::=\; [\cdot]
\;\mid\; \ectx\,;\,e
\;\mid\; \textsf{let}\;x = \ectx\;\textsf{in}\;e
\;\mid\; \ectx + e \;\mid\; v + \ectx
\;\mid\; \ectx == e \;\mid\; v == \ectx
\;\mid\; \textsf{if}\;\ectx\;\textsf{then}\;e\;\textsf{else}\;e
\;\mid\; \setc{l}{\ectx}.
\]
We write $\ehole{e}$ for the expression obtained by plugging $e$ into the hole. Thus
$E(t)=\ehole{\bbegin}$ asserts that the redex thread $t$ is about to execute is
$\bbegin$, sitting inside the surrounding continuation $\ectx$.

\subsection{The command rules in context form}

Every command rule now fires on its redex inside an arbitrary context $\ectx$,
preserving the continuation. The configuration is the six-tuple
$\langle\sigma,\lambda,\theta,E,\hat{\sigma},\hat{E}\rangle$ of
Section~\ref{subsec:semantics}.

\begin{itemize}
\item $E(i)=\ehole{\bbegin}$ and $E'=E[i\mapsto \ehole{\unit}]$: the $\bbegin$ reduces to
$\unit$ \emph{in place}, leaving the continuation $\ectx$.
\item $\hat{E}'=\ehole{\unit}$: the saved resume point is $\ehole{\unit}$,
the body \emph{after} $\bbegin$, which is precisely what $\textsc{Retry}$
reinstates.
\end{itemize}
\textsc{Retry} rolls the store back to
the snapshot $\hat{\sigma}$, releases the lock ($\lambda = \bot$), and
re-arms $E(i)$ as $\bbegin\,;\,\hat{E}$ so that the thread must re-acquire the
lock through $\textsc{Begin}$ before re-running.

\subsection{Administrative reductions}
\label{subsec:admin}

\[
\begin{aligned}
\ehole{\textsf{let}\;x = v\;\textsf{in}\;e} &\;\reduces\; \ehole{\subst{e}{x}{v}} &
\ehole{\unit \,;\, e} &\;\reduces\; \ehole{e} \\
\ehole{n_1 + n_2} &\;\reduces\; \ehole{n} \quad (n = n_1 + n_2) &
\ehole{\textsf{if}\;\textsf{true}\;\textsf{then}\;e_1\;\textsf{else}\;e_2}
                      &\;\reduces\; \ehole{e_1} \\
  \ehole{\textsf{if}\;\textsf{false}\;\textsf{then}\;e_1\;\textsf{else}\;e_2}
  & \;\reduces\; \ehole{e_{2}}
\end{aligned}
\]
Here the notation is substitution $\subst{e}{x}{v}$ in the \textsf{let}
rather than map update.




%% =====================================================================
\section{Initial configuration and termination}
\label{sec:init-final}

The transition system $\mathcal{S}=(\Phi,\rightarrow)$ has an \emph{infinite}
state space: a configuration $\langle\sigma,\lambda,\theta,E,\hat{\sigma},\hat{E}\rangle$
ranges over all stores, all thread maps, all snapshots, and all saved bodies, and
there are infinitely many of each. A program does not fix a single starting point;
it determines a \emph{set} of initial configurations, and a run may begin from any
one of them.

\subsection{Initial configurations}

The set of \emph{initial configurations} is
\[
\mathcal{I} \;=\;
\bigl\{\, \langle\, \sigma_0,\ \bot,\ \bot,\ E_0,\ \sigma_0,\ \unit \,\rangle
\;:\; \sigma_0, E_0 \text{ meet the conditions below} \,\bigr\}
\;\subseteq\; \Phi ,
\]

and a run may start from any $C_0 \in \mathcal{I}$.
\begin{itemize}
\item \emph{Locations are fixed and finite.} $\Loc$ is a fixed finite set and the
initial store maps every location to zero: $\sigma_0 : \Loc \rightarrow \mathbb{N}$
with $\forall l \in \Loc.\ \sigma_0(l) = 0$. So $\sigma_0$ is fixed, and
$\mathcal{I}$ varies only with  $\sigma_{0}$.
\item \emph{Threads are fixed and finite.} $dom(E_0) = \ThId$ is a fixed finite set
of thread identifiers, and $E_0 : \ThId \rightarrow \Expr$ is the initial program,
mapping each thread to its starting expression. No thread is ever created or
destroyed, so $dom(E) = \ThId$ at every configuration.
\item \emph{Every thread starts with a real program.} $E_0(i) \neq \unit$ for each
$i \in \ThId$: a thread already at the unit value is finished and cannot step, so an
initial configuration must give every thread an unreduced expression for the system
to make progress.
\item \emph{Nothing is live yet.} $\lambda = \bot$ (no thread owns the lock) and
$\theta = \bot$ (no transaction is live), since before any $\bbegin$ no thread is in
a transaction. The snapshot and saved body are given values here,
$\hat{\sigma}_0 = \sigma_0$ and $\hat{E}_0 = \unit$, so the configuration is fully
defined from the outset.
\end{itemize}

Fixing any $C_0 \in \mathcal{I}$, the configurations that matter are those
reachable from it,
\[
\mathrm{Reach}(C_0) \;=\; \{\, C \;:\; C_0 \rightarrow^{*} C \,\},
\]
where $\rightarrow^{*}$ is the reflexive--transitive closure of $\rightarrow$.
There is no scheduler: the nondeterminism of $\rightarrow$ is inherent, and at each
configuration \emph{any} thread whose redex is enabled may take the next step.
These reachable step sequences are the program's \emph{executions}.




%% =====================================================================
\section{Worked example: well-formedness is structural}
\label{sec:examples}

Because $\lambda$ now names the \emph{owner thread} and every transactional rule
carries the premise $\lambda = i$, a thread can take a transactional step only
inside a transaction it has itself opened. Call a program \emph{well-formed} when
every $\getc{l}$, $\setc{l}{e}$, or $\retry$ it reaches sits inside a
matching enclosing $\bbegin\dots\eend$. The point of this section is that
well-formedness is \emph{structural}: an ill-formed redex simply has no rule to
fire, so the run gets stuck rather than performing the action. We never need
well-formedness as a separate side condition.

\paragraph{Conventions.}
There is a single shared location $l$; we write $\st{n}$ for the store with
$\sigma(l)=n$. The configuration is the six-tuple
$\langle\sigma,\lambda,\theta,E,\hat{\sigma},\hat{E}\rangle$ of
Section~\ref{subsec:semantics}, with $\lambda$ the owner thread and $\theta$ the live
transaction id. Thread identifiers are $1,2\in\ThId$; $E$ is shown as a map
$\emap{1{:}e_1,\,2{:}e_2}$. Every run starts from
$C_0=\langle \st{0},\ \bot,\ \bot,\ E_0,\ \st{0},\ \unit\rangle$.

\subsection{An ill-formed \texttt{get} cannot fire}

Let thread $1$ run the ill-formed program $P \equiv \getc{l}$ --- a read with no
enclosing $\bbegin$. The initial configuration is
\[
C_0 \;=\; \langle \st{0},\ \bot,\ \bot,\ \emap{1{:}\getc{l}},\ \st{0},\ \unit\rangle .
\]
The active thread is $1$ and its redex is $\getc{l}$ (context $\ectx=[\cdot]$).
We check every rule of Figure~\ref{fig:semantics}:
\begin{itemize}
\item $\textsc{Get}$ requires $\lambda = 1$, but $\lambda = \bot$. \emph{Fails.}
\item $\textsc{Begin}$ requires the redex to be $\bbegin\,;\,e$, but it is
$\getc{l}$. \emph{Fails.}
\item $\textsc{Set},\textsc{End},\textsc{Retry}$ require their own redexes
($\setc{l}{v}$, $\eend$, $\retry$) and $\lambda = 1$; both premises fail.
\end{itemize}
No rule matches, so $C_0$ takes \emph{no} step:
\[
\langle \st{0},\ \bot,\ \bot,\ \emap{1{:}\getc{l}},\ \st{0},\ \unit\rangle
\ \not\rightarrow .
\]
The ill-formed $\getc{l}$ never executes; the thread is stuck. Contrast the
well-formed version, $\bbegin\,;\,(\getc{l}\,;\,\eend)$. There the leading
$\bbegin$ fires first (setting $\lambda:=1$), and only \emph{then}, with
$\lambda = 1$, can the $\getc{l}$ fire: it is exactly the $\bbegin$ that grants
the ownership $\getc{l}$ demands.

\subsection{A thread cannot act inside another thread's transaction}

This is the case the premise $\lambda = i$ is really designed to rule out. Take
two threads,
\[
E_0 \;=\; \emap{1{:}\bbegin\,;\,(\getc{l}\,;\,\eend),\ 2{:}\getc{l}} ,
\]
so thread $1$ is well-formed and thread $2$ tries a bare $\getc{l}$. Let thread
$1$ open its transaction:
{\small
\[
\begin{aligned}
&\langle \st{0},\ \bot,\ \bot,\ E_0,\ \st{0},\ \unit\rangle\\
&\quad\xrightarrow{\ \bbegin\,@\,t_1\ }\
\langle \st{0},\ 1,\ t_1,\ \emap{1{:}\unit\,;\,(\getc{l}\,;\,\eend),\ 2{:}\getc{l}},\ \st{0},\ \getc{l}\,;\,\eend\rangle .
\end{aligned}
\]}%
Now $\lambda = 1$: thread $1$ owns the live transaction $t_1$. Suppose thread $2$
is the next to step; its redex is $\getc{l}$. The only rule that could
fire is $\textsc{Get}$, and it requires $\lambda = 2$ --- but $\lambda = 1$. The
premise fails, so thread $2$'s read is \emph{not enabled}:
\[
\text{there is no } C' \text{ with }\quad
\langle \st{0},\ 1,\ t_1,\ \emap{1{:}\dots,\ 2{:}\getc{l}},\ \st{0},\ \getc{l}\,;\,\eend\rangle
\ \xrightarrow{\ \getev{l}{0}{\,\cdot\,}\ }\ C' .
\]
Under the old rule (premise $\lambda = t$, with $t$ unconstrained by the acting
thread) this step \emph{would} have fired, letting thread $2$ read inside thread
$1$'s transaction and mislabel it $@t_1$. The premise $\lambda = i$ removes that:
only the owner may step, so thread $2$ must wait until thread $1$ commits
($\lambda$ returns to $\bot$) and then open its own transaction. Well-formedness
holds by construction.

%% =====================================================================
\section{Traces, Transactions, and Histories}
\label{sec:history}

Throughout, sets are written with capitals --- $E$ (the events of a transaction),
$H$ (a history), and the boldface $\mathbf{T}$ (the set of \emph{all} transactions
of a trace). We range over transaction identifiers with $t_1, t_2, \dots$. A
\emph{single} transaction is written with a plain capital $T$ (and $S$ when we need
a second one to compare against, as in arbitration below); $T_{t}$ names the
transaction whose identifier is $t$. The typographic contrast is deliberate: the
boldface $\mathbf{T}$ is always the whole set, whereas $T$, $S$, and $T_{t}$ are
individual transactions drawn from it. The examples all refer to one running trace,
\[
\tau \;=\; \bbegin\,@\,t_1 \cdot \getev{l}{0}{t_1} \cdot \ladm[,1] \cdot \ladm[,2] \cdot
\setev{l}{1}{t_1} \cdot \eend\,@\,t_1 \cdot \bbegin\,@\,t_2 \cdot \getev{l}{1}{t_2}
\cdot \retry\,@\,t_2 .
\]

Here $\ladm[,1]$ and $\ladm[,2]$ are \emph{administrative steps}: internal reductions the
running transaction takes while evaluating its body (the $\adm$-steps of
Section~\ref{subsec:admin} --- a $\textsf{let}$-substitution, an arithmetic step, a
conditional). The Trace definition below makes this precise.

\subsection{Trace}
\begin{definition}[Trace]
The \emph{trace} of a run $C_0 \xrightarrow{\ell_1} C_1 \xrightarrow{\ell_2} \cdots
\xrightarrow{\ell_k} C_k$ is the sequence of the labels between the
configurations, $\ell_1\,\ell_2\cdots\ell_k$. Each label $\ell$ is either a
\emph{transition label} --- one of $\bbegin\,@\,t$, $\getev{l}{v}{t}$,
$\setev{l}{v}{t}$, $\eend\,@\,t$, $\retry\,@\,t$ --- or an \emph{administrative
step} $\ladm[,i]$ (Section~\ref{subsec:admin}), the label of an internal reduction the transaction takes while
evaluating its body (a $\adm$-step: a $\textsf{let}$-substitution, an arithmetic
step, a conditional). A run may take \emph{arbitrarily many} administrative steps
in a row, so they are indexed $\ladm[,1], \ladm[,2], \ladm[,3], \dots$
\end{definition}

\begin{example}
$\tau$ above is a trace. Between $\getev{l}{0}{t_1}$ and $\setev{l}{1}{t_1}$ it
takes \emph{two} administrative steps --- $\ladm[,1]$ (the $\textsf{let}$-substitution)
and $\ladm[,2]$ (the addition $0+1 \to 1$) --- and a longer transaction body would
take more.
\end{example}

\subsection{Transaction}
\begin{definition}[Transaction]
A \emph{transaction} is a pair $T = (E,\po)$. Every event in $E$ carries the
\emph{same} transaction identifier $t$ --- that shared tag is exactly what groups
these events into one transaction.
\begin{itemize}
\item $E$ is a set of pairs $(\iota, e)$ --- an identifier $\iota$ together with an
event $e$ tagged with the transaction. An event is one of $\bbegin\,@\,t$,
$\getev{l}{v}{t}$, $\setev{l}{v}{t}$, $\eend\,@\,t$, or $\retry\,@\,t$: the
transaction's opening, its reads and writes, and its terminator. (Administrative
steps $\ladm[,i]$ are trace labels, not events, so they are not members of $E$.)
\item $\po$ is a \emph{total order} on the events of $E$. Writing $a \prec b$ for
$(a,b) \in \po$, it is
  \begin{itemize}
  \item \emph{irreflexive}: no event precedes itself, $\neg(a \prec a)$;
  \item \emph{transitive}: $a \prec b$ and $b \prec c$ imply $a \prec c$;
  \item \emph{antisymmetric}: $a \prec b$ and $b \prec a$ imply $a = b$;
  \item \textbf{totality}: any two distinct events are comparable, $a \prec b \vee b \prec a$.
  \end{itemize}
\end{itemize}
\end{definition}

\begin{example}
Transaction $t_1$ of $\tau$ is the single transaction $T_{t_1} = (E,\po)$ with
\[
E = \{\, (\iota_0, \bbegin\,@\,t_1),\ (\iota_1, \getev{l}{0}{t_1}),\
         (\iota_2, \setev{l}{1}{t_1}),\ (\iota_3, \eend\,@\,t_1) \,\},
\]
ordered $(\iota_0,\_) \prec (\iota_1,\_) \prec (\iota_2,\_) \prec (\iota_3,\_)$ ---
begin, then the read, then the write, then end.
\end{example}



\subsection{Committed transaction}
For a total order $\po$ and a set of events $E$, we define
\[
\operatorname{max}_{\po}(E) =
\begin{cases}
e, & \text{if } e \in E \text{ and } \forall e' \in E.\; e = e' \lor (e',e) \in \po,\\[0.8ex]
\text{undefined}, & \text{if } E = \emptyset.
\end{cases}
\]

\[
\operatorname{min}_{\po}(E) =
\begin{cases}
e, & \text{if } e \in E \text{ and } \forall e' \in E.\; e = e' \lor (e,e') \in \po,\\[0.8ex]
\text{undefined}, & \text{if } E = \emptyset.
\end{cases}
\]

\begin{definition}[Committed transaction]
A transaction $T$ is \emph{committed} when its least event is a begin and its
greatest event is an end:
\[
\operatorname{min}_{po}(fst(T))\ \text{is a}\ \bbegin\,@\,t \qquad\text{and}\qquad
\operatorname{max}_{po}(fst(T))\ \text{is an}\ \eend\,@\,t .
\]
Because $\min_{po}$ and $\max_{po}$ are the extremes of $\po$, this forces $\bbegin\,@\,t$ to
precede $\eend\,@\,t$ and every event of $T$ to lie between them. For a trace with
transaction set $\mathbf{T}$, we write $\mathrm{Committed}(\mathbf{T}) \subseteq
\mathbf{T}$ for its committed transactions.
\end{definition}

\begin{example}
For $t_1$ above, $\operatorname{min}_{po}(fst(T_{t_1})) = (\iota_0, \bbegin\,@\,t_1)$ and
$\operatorname{max}_{po}(fst(T_{t_1})) = (\iota_3, \eend\,@\,t_1)$: the least event is a begin and the
greatest an end, so $T_{t_1}$ is committed.
\end{example}

\subsection{Uncommitted transaction}
\begin{definition}[Uncommitted transaction]
A transaction $T=(E,\po)$ is \emph{uncommitted} if its maximal event
$\operatorname{max}_{po}(fst(T))$ is not an $\eend\,@\,t$ event. Equivalently, $T$ does not have
a matching end event (e.g., its maximal event is a $\retry\,@\,t$, or it
has no end event at all).

For a trace $\tau$ with transaction set $\mathbf{T}$, the set of uncommitted
transactions is
\[
\mathrm{Uncommitted}(\mathbf{T})
=
\mathbf{T}\setminus \mathrm{Committed}(\mathbf{T}).
\]
\end{definition}

\begin{example}
For $t_2$ above, $\max(T_{t_2})$ is $\retry\,@\,t_2$, not an $\eend\,@\,t_2$, so
$T_{t_2} \in \mathrm{Uncommitted}(\mathbf{T}) = \mathbf{T} \setminus \mathrm{Committed}(\mathbf{T})$.
\end{example}


\subsection{Arbitration}
\label{subsec:arbitration}
\begin{definition}[Arbitration $\mathrm{AR}$]
Let $\mathbf{T}$ be the set of transactions of a trace. \emph{Arbitration} is a
binary relation $\mathrm{AR} \subseteq \mathbf{T} \times \mathbf{T}$ that is a total
order, defined by: for $S, T \in \mathbf{T}$,
\[
S <_{\mathrm{AR}} T \quad\text{iff}\quad S\text{'s terminator }
(\eend\,@\,S \text{ or } \retry\,@\,S)\ \text{precedes}\ \bbegin\,@\,T\ \text{in the trace.}
\]
\end{definition}

\begin{example}
In $\tau$, $\eend\,@\,t_1$ precedes $\bbegin\,@\,t_2$, so $T_{t_1} <_{\mathrm{AR}} T_{t_2}$.
\end{example}

\subsection{History}
\begin{definition}[History]
A \emph{history} is a pair
\[
H \;=\; (\mathbf{T},\ \mathrm{AR}),
\]
whose first component $\fst(H) = \mathbf{T} = \{\, T_1, \dots, T_n \,\} = \{\, (E_1,\po_1), \dots, (E_n,\po_n) \,\}$
is the (finite) set of all transactions of a trace --- committed and uncommitted ---
and whose second component $\snd(H) = \mathrm{AR} \subseteq \mathbf{T} \times \mathbf{T}$
is the arbitration total order on them (Section~\ref{subsec:arbitration}).
\end{definition}

\begin{example}
The trace $\tau$ yields the history
\[
H = (\mathbf{T},\ \mathrm{AR}), \qquad
\mathbf{T} = \fst(H) = \{\, T_{t_1},\ T_{t_2} \,\}, \qquad
\mathrm{AR} = \snd(H) = \{\, (T_{t_1}, T_{t_2}) \,\},
\]
i.e.\ $T_{t_1} <_{\mathrm{AR}} T_{t_2}$, with $T_{t_1}$ committed and $T_{t_2}$
uncommitted.
\end{example}


\subsection{Local (pure) reductions}
\label{subsec:local}
Addition, \code{let}, and \code{if} never touch the shared store or
transaction state; they rewrite only the expression of the scheduled
thread. We give them as a separate one-step relation
\(e \rightsquigarrow e'\) on \emph{redexes}, independent of the
six-tuple:

\begin{mathpar}
\inferrule*[left=\textsc{Add}]{
  n = n_1 + n_2
}{
  n_1 + n_2 \rightsquigarrow n
}
\and
\inferrule*[left=\textsc{Let}]{
}{
  \textsf{let}\;x = v\;\textsf{in}\;e \rightsquigarrow e[v/x]
}
\and
\inferrule*[left=\textsc{IfTrue}]{
}{
  \textsf{if}\;\textsf{true}\;\textsf{then}\;e_1\;\textsf{else}\;e_2 \rightsquigarrow e_1
}
\and
\inferrule*[left=\textsc{IfFalse}]{
}{
  \textsf{if}\;\textsf{false}\;\textsf{then}\;e_1\;\textsf{else}\;e_2 \rightsquigarrow e_2
}
\end{mathpar}

\subsection{Lifting local reductions into the global transition system}
\label{subsec:lift}

To connect \(\rightsquigarrow\) with the labeled transition relation
\(\rightarrow\) on configurations, we add one more rule that lets any
thread take an administrative step, labeled \(\ladm\), whenever its
current expression's redex is a pure one:

\begin{mathpar}
\inferrule*[left=\textsc{Local}]{
  E(i)=\ehole{r} \\ r \rightsquigarrow e
}{
  \langle\sigma,\lambda,\theta,E,\hat{\sigma},\hat{E}\rangle
  \xrightarrow{\ladm}
  \langle\sigma,\lambda,\theta,E[i\mapsto\ehole{e}],\hat{\sigma},\hat{E}\rangle
}
\end{mathpar}
\textsc{Local} leaves \(\sigma\), \(\lambda\), \(\theta\), \(\hat{\sigma}\), and
\(\hat{E}\) untouched. Together, \textsc{Local} plus the five rules of
Figure~\ref{fig:semantics} give a reduction for every production of the
expression grammar.

%% =====================================================================
\section{Consistency}
\label{sec:consistency}

The declarative model of Section~\ref{sec:history} lets us say when a history
$H = (\mathbf{T}, \mathrm{AR})$ is \emph{consistent}, following Cerone et al. The
axioms below range over its transactions $\fst(H) = \mathbf{T}$, using its
arbitration order $\snd(H) = \mathrm{AR}$. Two axioms do the work: \textsc{Int}
constrains a get against the sets of its \emph{own} transaction, and
\textsc{Ext} constrains a get against the sets of \emph{other}, committed,
transactions.

\subsection{Internal consistency}
\label{sec:int}

For a history $H = (\mathbf{T}, \mathrm{AR})$, Cerone et al.'s internal-consistency
axiom is
\[
\forall (E, \po) \in \fst(H).\ \forall e \in E.\ \forall x, n.\quad
\bigl(e = (\_,\, \rdev{x}{n}) \,\wedge\, \po^{-1}(e) \cap \HEvent_x \neq \emptyset\bigr)
\implies
\maxpo\bigl(\po^{-1}(e) \cap \HEvent_x\bigr) = (\_,\, \anyop{x}{n}),
\tag{\textsc{Int}}
\]
where $\HEvent_x$ is the set of get and set events on $x$, and
$(\_,\anyop{x}{n})$ is a get or set of $n$ on $x$. It ranges over \emph{every}
transaction of the trace, committed or not: within one transaction a get of $x$
must return the value of the $\po$-latest earlier access to $x$.

\begin{example}[$\textsc{Int}$ needs no commit]
Neither transaction below commits, yet $\textsc{Int}$ still pins both gets. Take
the trace
\[
\tau:\quad \bbegin\,@\,t_1 \cdot \setev{x}{2}{t_1} \cdot \getev{x}{2}{t_1} \cdot \retry\,@\,t_1
      \cdot \bbegin\,@\,t_2 \cdot \setev{x}{5}{t_2} \cdot \getev{x}{5}{t_2}
\]
($t_1$ aborts, $t_2$ is left pending). For $\getev{x}{2}{t_1}$,
$\po^{-1}(e)\cap\HEvent_x = \{\setev{x}{2}{t_1}\}$, whose $\maxpo$ is a set of $2$
on $x$, so the get is forced to $2$; likewise $\getev{x}{5}{t_2}$ is forced to
$5$. Committedness never entered.
\end{example}

\begin{example}[external get: $\textsc{Int}$ vacuous]
When a get is a transaction's first access to $x$, $\textsc{Int}$ is silent. In
\[
\tau:\quad \bbegin\,@\,t_1 \cdot \setev{x}{7}{t_1} \cdot \eend\,@\,t_1
      \cdot \bbegin\,@\,t_2 \cdot \getev{x}{7}{t_2}
\]
($t_1$ committed), the get $\getev{x}{7}{t_2}$ has
$\po^{-1}(e)\cap\HEvent_x = \emptyset$, so the premise fails and $\textsc{Int}$
holds \emph{vacuously}.
\end{example}

\subsection{External consistency}
\label{sec:ext}

Internal consistency constrains a get only when the same transaction has already
touched that location --- when $\po^{-1}(e) \cap \HEvent_x \neq \emptyset$. Every
other get is one whose value \emph{cannot} come from within its own transaction: a
get of $x$ not preceded, in its transaction, by any access to $x$. We call such a
get \emph{external}, because the value it returns must come from \emph{another}
transaction. External gets are exactly the ones $\textsc{Int}$ leaves free --- its
premise is vacuous, as in the last example --- and external consistency is what
fixes them.

Following Cerone, we summarise a transaction's effect on a location $x$ with two
judgements. $T \vdash \mathsf{Set}\ x : n$ holds when $T$ sets $x$ and its
$\po$-last set of $x$ has value $n$ --- the value $T$ leaves in $x$. And
$T \vdash \mathsf{Get}\ x : n$ holds when $T$'s $\po$-first access to $x$ is a get
returning $n$ --- precisely an \emph{external} get of $x$ by $T$.

The value such a get may return comes from the committed transactions that took
effect before it, in the arbitration order $\mathrm{AR}$ of
Section~\ref{subsec:arbitration}. Let $\mathrm{AR}^{-1}(T) = \{\, S \mid S <_{\mathrm{AR}} T \,\}$
be the transactions arbitrated before $T$. Since $\mathrm{AR}$ orders \emph{all}
transactions --- committed and aborted alike --- but an aborted set must never be
observed, the candidate setters $S$ have to be restricted to \emph{committed}
transactions. We make that restriction explicit in the axiom below, as the
$S \in \mathrm{Committed}(\fst(H))$ conjunct of the setter set $W_x$.

We do \emph{not} model a separate visibility relation. $\mathrm{VIS}$ would record
that one transaction can \emph{see} another's effects --- a distinct notion only
under weaker isolation, where serializability fails, visibility is strictly weaker
than arbitration ($\mathrm{VIS} \subsetneq \mathrm{AR}$), and $\mathrm{VIS}$ must be
given on its own. Here there is no such gap: every committed transaction's updates
are available to every $\mathrm{AR}$-later transaction, so arbitration alone
suffices.

External consistency then says: an external get returns the value set by the
$\mathrm{AR}$-latest committed setter of that location, or the initial value $0$ if
there is none:
\[
\begin{aligned}
&\forall T \in \fst(H).\ \forall x, n.\quad T \vdash \mathsf{Get}\ x : n \implies {}\\
&\quad \bigl(\mathrm{AR}^{-1}(T) \cap W_x = \emptyset \,\wedge\, n = 0\bigr)\ \vee\
       \mathrm{max}_{\mathrm{AR}}\bigl(\mathrm{AR}^{-1}(T) \cap W_x\bigr) \vdash \mathsf{Set}\ x : n,\\[2pt]
&\text{where}\quad W_x \;=\; \{\, S \mid S \in \mathrm{Committed}(\fst(H))\ \wedge\ S \vdash \mathsf{Set}\ x : \_ \,\}.
\end{aligned}
\tag{\textsc{Ext}}
\]

\begin{example}[committed setter is visible]
$t_1$ commits its set, so $\textsc{Ext}$ pins $t_2$'s get to it. In
\[
\tau:\quad \bbegin\,@\,t_1 \cdot \setev{x}{7}{t_1} \cdot \eend\,@\,t_1
      \cdot \bbegin\,@\,t_2 \cdot \getev{x}{7}{t_2}
\]
$\eend\,@\,t_1$ precedes $\bbegin\,@\,t_2$, so $T_{t_1} <_{\mathrm{AR}} T_{t_2}$, and
$T_{t_1}$ is committed with $T_{t_1} \vdash \mathsf{Set}\ x : 7$; thus
$T_{t_1} \in \mathrm{AR}^{-1}(T_{t_2}) \cap W_x$ and
$\mathrm{max}_{\mathrm{AR}}\bigl(\mathrm{AR}^{-1}(T_{t_2}) \cap W_x\bigr) = T_{t_1}$,
so $\textsc{Ext}$ forces the get to $7$.
\end{example}

\begin{example}[no committed setter: get is $0$]
$t_1$ aborts, so its set is uncommitted; $\textsc{Ext}$ still holds --- with the
initial value. In
\[
\tau:\quad \bbegin\,@\,t_1 \cdot \setev{x}{2}{t_1} \cdot \retry\,@\,t_1
      \cdot \bbegin\,@\,t_2 \cdot \getev{x}{0}{t_2}
\]
we have $T_{t_1} <_{\mathrm{AR}} T_{t_2}$, but $T_{t_1}$ is \emph{not} committed, so
$T_{t_1} \notin W_x$; hence $\mathrm{AR}^{-1}(T_{t_2}) \cap W_x = \emptyset$, and
$\textsc{Ext}$ forces the get to $0$ --- matching the rolled-back store.
\end{example}

\end{document}
